*DCVtestkit*: an R package for analysis of quality control dilution curves

### 2022-10-13 Manuscript/manuscript.qmd
Update manuscript with the changes recommended by reviewer.

`DCVtestkit` is available on GitHub <https://github.com/SLINGhub/DCVtestkit>. The documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>

@@

To address the issues mentioned above, we have created an R package, DCVtestkit, to assist analysts, not only to plot these dilution curves for many molecules easily but also providing additional statistical results, other than the Pearson Correlation Coefficient $R$ or coefficient of determination $R^2$, for these curves. These alternative statistical results may be helpful in better describing the shape of the dilution curve. The R package also provides several ways of reporting the dilution data, including an interactive html file. This allows the analyst to group, filter and sort the dilution plots, to only look at the relevant ones, such as the saturated dilution curves, and identify problematic molecules quickly.

@@

Here, we introduce the workflow in which R package, DCVtestkit.

![DCVtestkit Workflow](images/README-Workflow.png)

@@

To address the paucity of parameters describing the goodness-of-fit of linear dilution curves, a few additional parameters are added in the R package `DCVtestkit`, on top of the popular Pearson Correlation Coefficient $R$ on a linear regression fit. Currently, the additional parameters are statistical results from Mandel's Fitting Test in @andrade2013 and the Percent Residual Accuracy in @logue2018. The software also fits a quadratic model on the dilution curve and calculates the concavity of the quadratic model to identify to verify if the dilution curve is predominantly non-linear at high or low concentrations.

Given an example linear curve, saturated curve and a curve with limit of detection (LOD) and their corresponding summary statistics, it can be observed that the corresponding Pearson Correlation Coefficient `r_corr` are really high $\ge0.9$ even for the curves that are non-linear. On the other hand, both the saturated and LOD curves give a much lower Mandel's Fitting Test p values `mandel_p_val` and Percent Residual Accuracy `pra_linear` values than the linear curve.

```{r Motivation, echo = FALSE}
linear_data <- data.frame(
  conc_var = c(10, 25, 40, 50, 60,
               75, 80, 100, 125, 150),
  signal_var = c(25463, 63387, 90624, 131274, 138069,
                 205353, 202407, 260205, 292257, 367924)
)

@@

tabular_display <- tibble::tibble(
  Concentration = c(10, 25, 40, 50, 60,
                    75, 80, 100, 125, 150), 
  `Linear\nsignal` = c(25463, 63387, 90624, 131274, 138069,
                    205353, 202407, 260205, 292257, 367924),
  `Saturated\nsignal` = c(5192648, 16594991, 32507833, 46499896,
                       55388856, 62505210, 62778078, 72158161,
                       78044338, 86158414),
  `Lod\nsignal` = c(500, 903, 1267, 2031, 2100,
                 3563, 4500, 5300, 8500, 10430)
)

@@

```{r create linear plot, echo = FALSE}
# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  linear_data, 
  "conc_var", 
  "signal_var")

@@

```{r create saturation plot, echo = FALSE}
# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  saturated_data, 
  "conc_var", 
  "signal_var")

@@

```{r create lod plot, echo = FALSE}
# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  lod_data, 
  "conc_var", 
  "signal_var")

@@

```{r ViewMotivationData1, echo = FALSE, fig.height=9, fig.width=13, fig.cap = "Plots and statistical results of three curves: Linear curve, Saturated curve and a curve with Limit Of Detection (LOD)"}
patchwork::wrap_plots(concentration_table, linear_plot, saturation_plot, lod_plot, ncol = 2, nrow = 2)
```

The Mandel's Fitting Test and Percent Residual Accuracy findings are also in agreement with the results from @vanloco2002 and @logue2018 respectively, highlighting that these two alternative parameters are more sensitive and advantageous than the Pearson Correlation Coefficient in their calibration curves and simulated experiments.

@@

`DCVtestkit` uses the function `evaluate_linearity` to group the dilution curves. At this moment, two workflows are proposed.

![Dilution Curve Grouping Workflow](images/README-LinearEvaluation.png)

While the threshold values given for Pearson Correlation Coefficient and Percent Residual Accuracy are based on @y.h.chan2003 and @logue2018 interpretation respectfully, it remains subjective and arbitrary. Moreover, it is also possible for the dilution curve to exhibit a sigmoidal relationship, or non-linear at both low and high concentrations, which the proposed workflow is unable to identify.

Nevertheless, while the proposed workflow is not infallible, `DCVtestkit` allows the analyst to change and optimise these threshold values to suit the congruency of the analyst's opinion about linearity and saturation of dilution curves. More information on how to perform such customisation can be found in the program documentation titled: ["Customise Linearity Evaluation Workflow"](https://slinghub.github.io/DCVtestkit/articles/lineval-customization.html) Furthermore, it is also possible to output these statistical summaries in Excel without the classification results.

@@

Lastly, `DCVtestkit` is able export the statistical results of each dilution curve in Excel. Furthermore, it can report the list of dilution plots as static charts with statistical summary results on the side in a single pdf file.

As some may find a pdf file to be too complex for further interpretation, the R package also provides an interactive viewer as a HTML folder to look at dilution plots in a trellis display framework, as introduced in @becker1996. Clicking on the `index.html` file inside the folder will open the interactive viewer that allows the analysts to group, filter and sort these trellis charts in a meaningful way based using @hafen2013 definition of "cognostics" or computer guiding diagnostics. This provides the analysts to be able to do exploratory data analysis such as identifying groups of molecules with linearity issues or finding out the effects of changing the Pearson Correlation Coefficient to another value. Such tasks can be challenging and time consuming if only the Excel and pdf files full of static plots are provided.

![Viewing Several Interactive Dilution Curves](images/README-TrellisOutput.png)

An example of an interactive viewer created by `DCVtestkit` can be accessed from <https://dcvtestkit-interactive-example.netlify.app/>. Such interactive visualisation was also utilised by @hafen2019 in the exploration of proteomics, metabolomics, and lipidomics data in a multi-omics fashion. The HTML folder can be compressed as a zip file and distributed to others like a pdf or Excel file. This interactive viewer can also be displayed in as a report created from Quarto (https://quarto.org/), an open-source scientific and technical publishing system. The report can be viewed at <https://dcvtestkit-quarto-example.netlify.app/>

R script and result outputs can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>

@@

A linear response in quality control dilution curves is one of the many checks used in metabolomics and lipidomics to assess the suitability of the molecules measured for further analysis. Despite its importance in providing high quality data, the shortage of software tools to analyse many dilution curves effectively and efficiently prevents researchers from doing it. An R package called `DCVtestkit` is created to rectify this issue by allowing many dilution curves to be analysed quickly and reliably. This is done by the use of alternative statistics, other than the Pearson Correlation Coefficient, to better understand the linearity of the dilution curves as well as reporting of statistical results and dilution plots that can be distributed to others. It is also able to provide an interactive trellis plot for exploratory data analysis. The R package is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>. As `DCVtestkit` is an opensource software, it is accessible to collaboration works which can help improve its usability and performance in the future.

>>>>

*DCVtestkit* is available on GitHub <https://github.com/SLINGhub/DCVtestkit>. The documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>

@@

To address the issues mentioned above, we have created an R package, *DCVtestkit*, to assist analysts, not only to plot these dilution curves for many molecules easily but also providing additional statistical results, other than the Pearson Correlation Coefficient $R$ or coefficient of determination $R^2$, for these curves. These alternative statistical results may be helpful in better describing the shape of the dilution curve. The R package also provides several ways of reporting the dilution data, including an interactive html file. This allows the analyst to group, filter and sort the dilution plots, to only look at the relevant ones, such as the saturated dilution curves, and identify problematic molecules quickly.

@@

Here, we introduce the workflow used by our R package, *DCVtestkit* in @fig-DCVtestkit-Workflow.

::: {#fig-DCVtestkit-Workflow layout-ncol=1}

![](images/README-Workflow.png)

*DCVtestkit* Workflow
:::

@@

@fig-curves gives a plot of three manually curated curves together with their corresponding summary statistics. The characteristic of three curves are as follows: One is a linear curve. Another is having a plateau at higher concentrations, also known as a saturated curve. The last example is a curve with a plateau at lower concentrations or experiencing a limit of detection (denoted as LOD curves in this paper). It can be observed that the corresponding Pearson Correlation Coefficient values, denoted as `r_corr`, are $\ge0.9$ ($0.99$, $0.95$ and $0.98$ respectively) even for the curves that are non-linear. These results are in agreement with @vanloco2002 as well as @logue2018 arguments that the Pearson Correlation Coefficient is a poor descriptor of non-linear calibration curves.

To improve the description of the goodness-of-fit for linear dilution curves, a few additional parameters, on top of the popular Pearson Correlation Coefficient $R$, are added in the R package *DCVtestkit*. The first additional parameters are statistical results from Mandel’s Fitting Test (as reported in @andrade2013) in which a low $p$ value gives sufficient evidence that a quadratic model is a better fit than a linear model, indicating the curve may not be linear. Another parameter calculated is the Percent Residual Accuracy (as reported in @logue2018) which ranges from $-\infty$ to $100$. If the curve is linear, the value should be close to $100$. Finally, the software also fits a quadratic model on the dilution curve and calculates the concavity, trying to verify if the dilution curve is predominantly non-linear at high (concavity $<0$) or low (concavity $>0$) concentrations.

Returning to @fig-curves, both the saturated and LOD curves give a much lower Mandel’s Fitting Test $p$ values, denoted as `mandel_p_val`, ($1.66 * 10^{-4}$ and $2.56 * 10^{-3}$ respectively vs $0.38$). Correspondingly, ranging from $-\infty$ to $100$, the Percent Residual Accuracy values, denoted as `pra_linear`, are much lower in the saturated and LOD curves compared to the linear curve ($62.30$ and $74.69$ respectively vs $94.32$).

```{r Motivation, echo = FALSE}
#| echo: false

linear_data <- data.frame(
  conc_var = c(10, 25, 40, 50, 60,
               75, 80, 100, 125, 150),
  signal_var = c(25463, 63387, 90624, 131274, 138069,
                 205353, 202407, 260205, 292257, 367924)
)

@@

tabular_display <- tibble::tibble(
  Concentration = c(10, 25, 40, 50, 60,
                    75, 80, 100, 125, 150), 
  `Linear signal` = c(25463, 63387, 90624, 131274, 138069,
                    205353, 202407, 260205, 292257, 367924),
  `Saturated signal` = c(5192648, 16594991, 32507833, 46499896,
                       55388856, 62505210, 62778078, 72158161,
                       78044338, 86158414),
  `LOD signal` = c(500, 903, 1267, 2031, 2100,
                 3563, 4500, 5300, 8500, 10430)
)

@@

```{r create linear plot}
#| echo: false

# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  linear_data, 
  "conc_var", 
  "signal_var")

@@

```{r create saturation plot}
#| echo: false

# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  saturated_data, 
  "conc_var", 
  "signal_var")

@@

```{r create lod plot}
#| echo: false

# Get dilution summary
statistical_summary <- DCVtestkit::summarise_dilution_data(
  lod_data, 
  "conc_var", 
  "signal_var")

@@

```{r ViewMotivationData1}
#| echo: false
#| label: fig-curves
#| fig-height: 9
#| fig-width: 13
#| fig-cap: "Plots and statistical results of three curves: Linear curve, Saturated curve and a curve with Limit Of Detection (LOD)"

patchwork::wrap_plots(
  concentration_table, 
  linear_plot, 
  saturation_plot, 
  lod_plot, 
  ncol = 2, 
  nrow = 2
)
```

@@

*DCVtestkit* uses the function `evaluate_linearity` to group the dilution curves. At this moment, two workflows are proposed.

::: {#fig-Dilution-Grouping-Workflow layout-ncol=1}

![](images/README-LinearEvaluation.png)

Dilution Curve Grouping Workflows
:::

While the threshold values given for Pearson Correlation Coefficient and Percent Residual Accuracy are based on the interpretation given by @y.h.chan2003 and @logue2018, respectively, it remains subjective and arbitrary. Moreover, it is also possible for the dilution curve to exhibit a sigmoidal relationship, or non-linear, at both low and high concentrations, which the proposed workflow would be unable to identify.

Nevertheless, while the proposed workflow is not a perfect solution, *DCVtestkit* allows the analyst to change and optimise the threshold values, to adjust the determinants of linearity and saturation of dilution curves according to the user’s needs. More information on how to perform such customisation can be found in the program documentation titled: ["Customise Linearity Evaluation Workflow"](https://slinghub.github.io/DCVtestkit/articles/lineval-customization.html). Furthermore, it is also possible to output these statistical summaries in Excel without the classification results.

@@

Lastly, *DCVtestkit* can export the statistical results of each dilution curve in Excel. Furthermore, it can report the list of dilution plots as static charts with the statistical summary results included in a single pdf file.

As some may find the final pdf file too complex for further interpretation, the R package also provides an interactive viewer, as a HTML folder, to look at dilution plots in a trellis display framework, as introduced in @becker1996. Clicking on the `index.html` file inside the folder will open the interactive viewer that allows the user to group, filter and sort these trellis charts in a meaningful way, based on @hafen2013's definition of "cognostics" or computer guiding diagnostics. This allows the user to do exploratory data analysis, such as identifying groups of molecules with linearity issues or finding out the effects of changing the Pearson Correlation Coefficient to another value. Such tasks can be challenging and time consuming if only the Excel and pdf files full of static plots were provided.

::: {#fig-Viewing-Interactive-Curves layout-ncol=1}

![](images/README-TrellisOutput.png)

Viewing Several Interactive Dilution Curves
:::

An example of an interactive viewer created by *DCVtestkit* can be accessed from <https://dcvtestkit-interactive-example.netlify.app/>. Such interactive visualisation was also utilised by @hafen2019 in the exploration of proteomics, metabolomics and lipidomics data in a multi-omics approach. The HTML folder can be compressed as a zip file and shared as a pdf or Excel file. This interactive viewer can also be displayed as a report created from Quarto (https://quarto.org/), an open-source scientific and technical publishing system. The report can be viewed at <https://dcvtestkit-quarto-example.netlify.app/>
R script and result outputs can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>

An example of an interactive viewer created by *DCVtestkit* can be accessed from <https://dcvtestkit-interactive-example.netlify.app/>. Such interactive visualisation was also utilised by @hafen2019 in the exploration of proteomics, metabolomics, and lipidomics data in a multi-omics fashion. The HTML folder can be compressed as a zip file and distributed to others like a pdf or Excel file. This interactive viewer can also be displayed in as a report created from Quarto (<https://quarto.org/>), an open-source scientific and technical publishing system. The report can be viewed at <https://dcvtestkit-quarto-example.netlify.app/>. R script and result outputs can be found in the Supplementary Material or by accessing this GitHub page <https://github.com/JauntyJJS/DCVtestkit_Output>.

@@

A linear response in quality control dilution curves is one of the requirements in metabolomics and lipidomics to assess the suitability of the molecules measured for further analysis. Despite its importance in testing data quality, the shortage of software tools to analyse dilution curves effectively and efficiently prevents researchers from fully analysing data. An R package called *DCVtestkit* is created to rectify this issue by allowing many dilution curves to be analysed quickly and reliably. This is done by the use of alternative statistics, other than the Pearson Correlation Coefficient, to better understand the linearity of the dilution curves, as well as reporting of statistical results and dilution plots that can be distributed to others. It is also able to provide an interactive trellis plot for exploratory data analysis. The R package is available on GitHub <https://github.com/SLINGhub/DCVtestkit> while the documentation and tutorials can be accessed from <https://slinghub.github.io/DCVtestkit/>. As *DCVtestkit* is an opensource software, it is accessible to collaboration works which can help improve its usability and performance in the future.
